---
title: "p8105_hw2_yg3096"
output: pdf_document
date: "2025-09-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(lubridate)
path <- "data/fivethirtyeight_datasets"
```

## Problem 1
```{r}
pols <- readr::read_csv(file.path(path, "pols-month.csv"))
# Clean pols-month
pols_clean <- pols %>%
              # After checking the table, format of mon is "YYYY-M-D"
              separate(mon, into = c("year", "month", "day")) %>%
              mutate(year = as.integer(year), month = as.integer(month),
                     month = month.name[month], # Replace with month names
                     # After checking the table, there are no null values
                     president = case_when(prez_gop == 1 ~ "gop",
                                           prez_dem == 1 ~ "dem")) %>%
              select(-prez_gop, -prez_dem, -day)
head(pols_clean)

snp  <- readr::read_csv(file.path(path, "snp.csv"))
# Clean snp
snp_clean <- snp %>%
             # M-D-YY format
             separate(date, into = c("month", "day", "year")) %>%
             mutate(year = as.integer(year), 
                    # Change 2-digit year format into 4-digit
                    year = if_else(year < 50, year + 2000L, year + 1900L),
                    month = as.integer(month), month = month.name[month]) %>%
             select(year, month, close)
head(snp_clean)

unemp <- readr::read_csv(file.path(path, "unemployment.csv"))
# Clean unemployment
unemp_long <- unemp %>%
              pivot_longer(cols = Jan:Dec, names_to = "month",
              values_to = "unemployment_rate") %>%
              mutate(year = as.integer(Year),
              # Rename Jan/Feb/... to January/February/... for consistency
              month = month.name[match(month, month.abb)]) %>%
              select(year, month, unemployment_rate)
head(unemp_long)

# Merge snp into pols
pols_snp <- pols_clean %>%
            left_join(snp_clean, by = c("year", "month"))

# Then merge into unemployment
merged_all <- pols_snp %>%
              left_join(unemp_long, by = c("year", "month"))

# Check result
dim(merged_all)
range(merged_all$year, na.rm = TRUE)
sum(is.na(merged_all$close)) # Check NA values in close column
sum(is.na(merged_all$unemployment_rate)) # Check NA values in unemployment
```
The three raw data contains the number of national politicians who are  
democratic or republican, Standard & Poor's stock market index (S&P) and rate of  
unemployment by month. After merging these data set by year and month, dimension  
of result is `r paste(dim(merged_all), collapse = " x ")`, year range is `range(merged_all$year, na.rm = TRUE)`.  
Key variables contain `r paste(names(merged_all)[1:8], collapse=", ")`.  

## Problem 2
```{r}


```