---
title: "p8105_hw2_yg3096"
output: pdf_document
date: "2025-09-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(readxl)
library(dplyr)
```

## Problem 1
```{r}
# Set data path
path <- "data/fivethirtyeight_datasets"
pols <- readr::read_csv(file.path(path, "pols-month.csv"), 
                        show_col_types = FALSE)
# Clean pols-month
pols_clean <- pols %>%
              # After checking the table, format of mon is "YYYY-M-D"
              separate(mon, into = c("year", "month", "day")) %>%
              mutate(year = as.integer(year), month = as.integer(month),
                     month = month.name[month], # Replace with month names
                     # After checking the table, there are no null values
                     president = case_when(prez_gop == 1 ~ "gop",
                                           prez_dem == 1 ~ "dem")) %>%
              select(-prez_gop, -prez_dem, -day)
head(pols_clean)

snp  <- readr::read_csv(file.path(path, "snp.csv"), show_col_types = FALSE)
# Clean snp
snp_clean <- snp %>%
             # M-D-YY format
             separate(date, into = c("month", "day", "year")) %>%
             mutate(year = as.integer(year), 
                    # Change 2-digit year format into 4-digit
                    year = if_else(year < 50, year + 2000L, year + 1900L),
                    month = as.integer(month), month = month.name[month]) %>%
             select(year, month, close)
head(snp_clean)

unemp <- readr::read_csv(file.path(path, "unemployment.csv"), 
                         show_col_types = FALSE)
# Clean unemployment
unemp_long <- unemp %>%
              pivot_longer(cols = Jan:Dec, names_to = "month",
              values_to = "unemployment_rate") %>%
              mutate(year = as.integer(Year),
              # Rename Jan/Feb/... to January/February/... for consistency
              month = month.name[match(month, month.abb)]) %>%
              select(year, month, unemployment_rate)
head(unemp_long)

# Merge snp into pols
pols_snp <- pols_clean %>%
            left_join(snp_clean, by = c("year", "month"))

# Then merge into unemployment
merged_all <- pols_snp %>%
              left_join(unemp_long, by = c("year", "month"))

# Check result
dim(merged_all)
range(merged_all$year, na.rm = TRUE)
sum(is.na(merged_all$close)) # Check NA values in close column
sum(is.na(merged_all$unemployment_rate)) # Check NA values in unemployment
```
The three raw data contains the number of national politicians who are  
democratic or republican, Standard & Poor's stock market index (S&P) and rate of  
unemployment by month. After merging these data set by year and month, dimension  
of result is `r paste(dim(merged_all), collapse = " x ")`, year range is `range(merged_all$year, na.rm = TRUE)`.  
Key variables include `r paste(names(merged_all)[1:8], collapse=", ")`.  

## Problem 2
```{r}
# Set data path
path <- "data/tw.xlsx"
# Read and clean Mr. Trash Wheel sheet, updated 
mr <- read_excel(path, sheet = "Mr. Trash Wheel", skip = 1) %>%
      janitor::clean_names() %>% # Clean names
      select(-starts_with("x")) %>% # Delete NA columns
      filter(!is.na(dumpster)) %>% # Omit non-data rows
      # Round number of spotrs balls to nearest integer
      mutate(year = as.integer(year), 
             sports_balls = as.integer(round(sports_balls, 0)),
             wheel = "Mr. Trash Wheel")

# Read and clean Professor Trash Wheel sheet
prof <- read_excel(path, sheet = "Professor Trash Wheel", skip = 1) %>%
        janitor::clean_names() %>% # Clean names
        select(-starts_with("x")) %>% # Delete NA columns
        filter(!is.na(dumpster)) %>% 
        # No sports balls column in this sheet
        mutate(year = as.integer(year), wheel = "Professor Trash Wheel")

# Read and clean Gwynnda Trash Wheel sheet
gwynnda <- read_excel(path, sheet = "Gwynnda Trash Wheel", skip = 1) %>%
           janitor::clean_names() %>% # Clean names
           select(-starts_with("x")) %>% # Delete NA columns
           filter(!is.na(dumpster)) %>% 
           # Also no sports balls column in this sheet
           mutate(year = as.integer(year), wheel = "Gwynnda Trash Wheel")

trash_all <- bind_rows(mr, prof, gwynnda)
```
After cleaning and combining the three Trash Wheel datasets, the result dataset  
contains `r nrow(trash_all)` observations. Key variables include `dumpster`  
(the dumpster identifier), `weight_tons` (weight of trash collected),  
`volume_cubic_yards` (volume of trash) and counts of specific items such as  
`plastic_bottles`, `polystyrene` and `cigarette_butts`. For Mr. Trash Wheel,  
the dataset also includes `sports_balls`, rounded to the nearest integer. Across   
all years of available data, Professor Trash Wheel collected a total of
`r sum(trash_all$weight_tons[trash_all$wheel == "Professor Trash Wheel"], na.rm = TRUE)`  
tons of trash. In June of 2022, Gwynnda collected
`r sum(trash_all$cigarette_butts[trash_all$wheel == "Gwynnda Trash Wheel" & trash_all$year == 2022 & trash_all$month == "June"], na.rm = TRUE)` cigarette butts.

## Problem 3
```{r}
# Set data path and read data
path <- "data/zillow_data"
zori <- readr::read_csv(file.path(path, 
                                  "Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv"), 
                        show_col_types = FALSE)
zori_long <- zori %>%
             rename(zip = RegionName) %>%
             mutate(zip = stringr::str_pad(as.character(zip), 5, pad = "0")) %>%
             pivot_longer(cols = matches("^\\d{4}-\\d{2}-\\d{2}$"), # YYYY-MM-DD
                          names_to = "date", values_to = "zori") %>%
             mutate(date = ymd(date), year = year(date), 
                    month = month(date, label = TRUE, abbr = FALSE, 
                                  locale = "C") %>% 
            as.character()) %>%
            select(zip, year, month, date, zori)
dim(zori_long)
head(zori_long)

zip <- readr::read_csv(file.path(path, "Zip Codes.csv"), show_col_types = FALSE)
zip_lookup <- zip %>%
              janitor::clean_names() %>% 
              transmute(zip = str_pad(as.character(zip_code), width = 5, 
                                      pad = "0"),
              county = county, neighborhood = neighborhood) %>%
              # Divide by borough name            
              mutate(borough = case_when(
              str_detect(county, regex("^bronx", ignore_case = TRUE)) ~ "Bronx",
              str_detect(county, regex("^kings", 
                                       ignore_case = TRUE)) ~ "Brooklyn",
              str_detect(county, regex("^new york", 
                                       ignore_case = TRUE)) ~ "Manhattan",
              str_detect(county, regex("^queens", 
                                       ignore_case = TRUE)) ~ "Queens",
              str_detect(county, regex("^richmond", 
                                       ignore_case = TRUE)) ~ "Staten Island",
              TRUE ~ NA_character_)) %>%
              select(zip, borough, neighborhood) %>%
              distinct()

# Check result
zip_lookup %>% summarize(n = n(), na_zip = sum(is.na(zip)),
                         na_borough = sum(is.na(borough)),
                         na_neigh = sum(is.na(neighborhood)))
# Keep distinct matches only
zip_lookup <- zip_lookup %>%
              distinct(zip, .keep_all = TRUE)
head(zip_lookup)

# Merge data
zori_full <- zori_long %>%
             left_join(zip_lookup, by = "zip") %>%
             relocate(zip, borough, neighborhood, year, month, date, zori)
# Basic information of result
n_obs <- nrow(zori_full)
n_zip <- n_distinct(zori_full$zip)
n_neigh <- n_distinct(zori_full$neighborhood, na.rm = TRUE)
list(n_obs = n_obs, n_zip = n_zip, n_neigh = n_neigh)

# Find zip in lookup but not in zori
zip_only_in_lookup <- anti_join(zip_lookup %>% 
                                select(zip) %>% 
                                distinct(), zori_long %>% 
                                select(zip) %>% 
                                distinct(), by = "zip")
# Check result
nrow(zip_only_in_lookup)
head(zip_only_in_lookup)

# get 2020-01 and 2021-01
zori_jan <- zori_full %>%
            filter(month == "January", year %in% c(2020, 2021)) %>%
            select(zip, borough, neighborhood, year, zori) %>%
            pivot_wider(names_from  = year, values_from = zori, 
                        names_prefix = "y") %>%
            mutate(drop_2020_2021 = y2021 - y2020)

# Find 10 zip with most drop
top10 <- zori_jan %>%
         filter(!is.na(y2020), !is.na(y2021)) %>%
         arrange(drop_2020_2021) %>%
         slice_head(n = 10) %>%
         select(zip, borough, neighborhood, y2020, y2021, drop_2020_2021)
# Show result
knitr::kable(top10, caption = "Largest ZORI drops from Jan 2020 to Jan 2021")
```
The final tidy dataset contains `r n_obs` observations across `r n_zip` ZIP codes   
and `r n_neigh` neighborhoods. While the borough variable is complete, approximately   
142 ZIP codes do not have neighborhood information in the lookup file.  

A total of `r nrow(zip_only_in_lookup)` ZIP codes appear in the lookup dataset but not in the Zillow  
ZORI data. They are likely to be PO Box–only ZIPs, industrial or commercial  
areas or lacking sufficient rental market coverage. Therefore these ZIPs are  
excluded from Zillow’s rental index.  

When comparing rental prices between January 2020 and January 2021, the 10 ZIP      
codes with the largest declines are shown in the table above. These declines    
occurred mainly in higher-rent neighborhoods. Notably, one ZIP code lacks  
neighborhood label in the lookup data, which explains the occasional NA values   
in the neighborhood column.  